A hazards approach to the biometric analysis of infant mortality

### 2023-04-04 Slides/slides.Rmd
updates from paper

---
title: "A hazards approach to the biometric analysis of infant mortality"
author:
- Göran Broström\inst{1}
- Tommy Bengtsson\inst{2}
institute:
- \inst{2}CED, Lund University (tommy.bengtsson@ekh.lu.se)
- \inst{1}CEDAR, Umeå University (gb@ehar.se, https://github.com/goranbrostrom)
output:
  beamer_presentation:
    theme: Warsaw
classoption: t
---

@@

*   Two categories: 
    *   Endogenous: *inherited*, *delivery*, etc.
    *   Exogenous: *accidental*, *infectious diseases*, etc.
*   How to differentiate between the two categories without infornation of causes of death?

Bourgeois-Pichat's Biometric model.

@@

1.  Endogenous deaths only occur during the neonatal period (0--28 days). 
2.  On a specific time scale, exogenous infant mortality is uniformly distributed.

@@

\tiny
\emph{Note:} $C$ is not part of B-P's original definition.

@@

*   Knodel \& Kintner (1977).
    *    Discusses effects of breastfeeding habits.
*   Wrigley (1977).
    *    Problems with data recording.
*   Lynch, Greenhouse \& Brändström (1998).
    *    Mentions "constant hazard".
    *    Suggest linear regression methods.
*   Bengtsson (1999).
    *    Data quality.
*   Manfredini (2004).
    *    Deals with effect of climate variation.

@@

@@

*   Good

>>>>

---
title: "A hazards approach to the biometric analysis of infant mortality"
author:
- Göran Broström\inst{1}
- Tommy Bengtsson\inst{2}
institute:
- \inst{2}CED, Lund University (tommy.bengtsson@ekh.lu.se)
- \inst{1}CEDAR, Umeå University (gb@ehar.se, https://github.com/goranbrostrom)
output:
  beamer_presentation:
#      includes: 
#         header: gb_header.txt
#         before_body: gb_beforebody.txt
      theme: Warsaw
classoption: t
header-includes:
- \newcommand{\emp}[1]{\textcolor{blue}{#1}}
---

@@

*   Two categories: 
    *   \emp{Endogenous}: *inherited*, *delivery*, etc.
    *   \emp{Exogenous}: *accidental*, *infectious diseases*, etc.
*   How to differentiate between the two categories without infornation of causes of death?

Bourgeois-Pichat's \emp{biometric model}.

@@

1.  \emp{Endogenous} deaths only occur during the \emp{neonatal} period (0--28 days). 
2.  On a specific time scale, \emp{exogenous} infant mortality is \emp{uniformly} distributed.

@@

\tiny
\emph{Note:} $C$ is \emp{not} part of B-P's original definition.

@@

*   Knodel \& Kintner (1977).
    *    Discusses effects of \emp{breastfeeding} habits.
*   Wrigley (1977).
    *    Problems with \emp{data recording}.
*   Lynch, Greenhouse \& Brändström (1998).
    *    Mentions \emp{"constant hazard"}.
    *    Suggest \emp{linear regression} methods.
*   Bengtsson (1999).
    *    Data \emp{quality}.
*   Manfredini (2004).
    *    Deals with effect of \emp{climate} variation.

@@

## Biometric analysis on the natural time scale

Given that the exogenous mortality follows an Exponential distribution on the log-cube scale, 
it follows that the \emp{cumulative hazards} function \emp{$H$} on the \emp{natural} time scale is

\begin{equation*}
H(t; \lambda) = \lambda \log^3(t + 1), \; t \ge 0,
\end{equation*}

and it follows that the \emp{hazard} function \emp{$h$} is given by

\begin{equation*}
h(t; \lambda) = \lambda \frac{3 \log^2(t + 1)}{t + 1}, \; t \ge 0,
\end{equation*}

As $\lambda$ varies, this constitutes a \emp{proportional hazards} family of distributions.

## Exogenous mortality on the normal time scale

```{r normal, fig.height = 4}
dg <- function(x) 3 * (log(1 + x))^2 / (1 + x) * 365 / (log(366))^3
oldpar <- par(mfrow = c(1, 2), las = 1)
x <- 0:365
y <- dg(x)
plot(x, y, type = "l", xlab = "Day", ylab = "Hazards",
     axes = FALSE, col = "blue", lwd = 2, main = "Hazard function")
axis(1, at = c(0, 28, 183, 365))
axis(2, las = 1, at = 0)
box()
abline(h = 0, v = 0)
abline(v = 28, lty = 2, col = "darkgreen")
y <- g(x)
plot(x, y, type = "l", xlab = "Day", ylab  = "Cum. hazards", main = "Cumulative hazards",
     col = "blue", axes = FALSE)
axis(1, at = c(0, 28, 183, 365))
axis(2, las = 1, at = 0)
abline(h = 0, v = 0)
box()
par(oldpar)
```

## Natural vs. log-cube time scales

\emp{Västerbotten} 1861--1890.

\includegraphics[width=\textwidth]{../paper2_files/figure-latex/natlogcube2-1.pdf}

## Natural time scale and Gompertz exogenous distribution

\emp{Skåne} 1711--1800

\includegraphics[width = \textwidth]{../paper2_files/figure-latex/escu-1.pdf}

@@

*   \emp{Advantages} of the hazards method.
       *    Fits naturally into modern survival analysis, allows \emp{censored} and \emp{truncated data}, and 
       *    The \emp{exogenous} component forms a class of \emp{proportional hazards} distributions.

*   \emp{Restrictions} for both models.
    *   Models for \emp{populations}, not for individuals.
    *   \emp{Less useful} for traditional \emp{survival analysis}.
    *   The \emp{log-cube} transform is \emp{not} universal.

*   \emp{Suggestions}.
    *   Work on the \emp{natural} time scale, if access to a computer.
    *   Try \emp{standard} distributions for \emp{exogenous}  mortality. 
    *   \emp{Remember} that we are dealing with \emp{competing risks}!
        *   Requires special care.
